---
title: "Data Cleaning Script: Chicago"
author: "Emmett McKinney"
date: "5/27/2019"
output: html_document
---

**THIS DOCUMENT JOINS SHUTOFF WITH CENSUS DATA & DOES SOME INITIAL ANALYSIS**

*Step 0: Intro*

To replicate this for other cities:

1. Find and replace all names of "Chicago" with the right city name
2. Update data cleaning strategy in Step 4.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/emmettmckinney/Dropbox (MIT)/RAship/ESI_Water_Affordability_031120/Data/WBEZ_Data/Chicago")
```

```{r}
#Install needed packages 

library(dplyr) # package to manage data
library(tidyr)
library(tidycensus) # package to census data
library(ggplot2) 
library(censusapi) 
library(tidyverse)
library(Hmisc)
library(readxl)
library(stargazer) #package to format tables
library(sf) # package to handle spatial objects 
```

```{r}
#install new census key, over-writing any old keys.

Sys.setenv(CENSUS_KEY="4c8277e5003e698cf77bbda813f1bcc7eb332b31") #set API key
readRenviron("~/.Renviron") #update R to use census key.
```

*Step 1. Read In Census Data* 

```{r}
geo <- get_acs( # get data for 2018
  geography = "zcta", #zip code tabulation area
  survey = "acs5",#ACS5 year survey
  cache_table = TRUE,  
  variables = "B02001_001",
  year = 2017,
  geometry = TRUE,
  output = "wide",
  ) %>% 
  filter(startsWith(GEOID, "60")) %>%
  select(GEOID, geometry) #only need GEOID and geometry columns
geo$GEOID <- as.numeric(geo$GEOID) 
```

```{r}
vars <- c(Total_Pop = "B02001_001", #total population
          White = "B02001_002", #white population
          Black = "B02001_003", 
          Native = "B02001_004", 
          Asian = "B02001_005", 
          Latino = "B03002_012",
          HIPI = "B02001_006", 
          Other = "B02001_007", 
          TWOPLUS = "B02001_008", 
          Utils_Incl = "B25069_001", #utilities ARE included in rent
          MHI = "B19013_001", #median household income
          TotalHH = "B25003_001", #total households
          TotalOwn = "B25003_002", #owner occupied households
          TotalRent = "B25003_003",
          WhiteOwn = "B25003A_002", #white owner occupied households 
          WhiteRent = "B25003A_003",
          BlackOwn = "B25003B_002", #black owner occupied households 
          BlackRent = "B25003B_003",
          HispOwn = "B25003I_002",
          HispRent = "B25003I_003",
          MedYrBlt = "B25037_001", #median age of structure, total
          RenYrBlt = "B25037_003", #median age of structure, renters
          OwnYrBlt = "B25037_002") #median age of structure, owners
```

*Step 2. Read In & Clean Census Data* 
```{r}
data_17 <- get_acs( # get data for 2017
  geography = "zcta", #zip code tabulation area
  survey = "acs5",#ACS5 year survey
  cache_table = TRUE,  
  variables = vars, 
  year = 2017
  ) %>% 
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) #zip codes in Chicago

data_est_17 <- data_17 %>% select(c(GEOID, contains("estimate"))) #split up by contains "estimate"

names(data_est_17) <- substring(names(data_est_17), 10) # renamed all the columns, dropping the first 10 letters
data_est_17 <- rename(data_est_17, GEOID = 1) #rename first column in df to say GEOID 
data_est_17$GEOID <- as.numeric(data_est_17$GEOID) #specify that this is numeric so that I can join it to numeric value (which will be "zip" in the shutoffs dataset)

##SAME THING FOR 2016

data_16 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2016
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) 

data_est_16 <- data_16 %>% select(c(GEOID, contains("estimate")))

names(data_est_16) <- substring(names(data_est_16), 10)
data_est_16 <- rename(data_est_16, GEOID = 1) 
data_est_16$GEOID <- as.numeric(data_est_16$GEOID)

data_15 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2015
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) 

data_est_15 <- data_15 %>% select(c(GEOID, contains("estimate")))

names(data_est_15) <- substring(names(data_est_15), 10)
data_est_15 <- rename(data_est_15, GEOID = 1) 
data_est_15$GEOID <- as.numeric(data_est_15$GEOID)

data_14 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2014
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) 

data_est_14 <- data_14 %>% select(c(GEOID, contains("estimate")))

names(data_est_14) <- substring(names(data_est_14), 10)
data_est_14 <- rename(data_est_14, GEOID = 1) 
data_est_14$GEOID <- as.numeric(data_est_14$GEOID)

data_13 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2013
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe))

data_est_13 <- data_13 %>% select(c(GEOID, contains("estimate")))

names(data_est_13) <- substring(names(data_est_13), 10)
data_est_13 <- rename(data_est_13, GEOID = 1) 
data_est_13$GEOID <- as.numeric(data_est_13$GEOID)

data_12 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2012
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) 

data_est_12 <- data_12 %>% select(c(GEOID, contains("estimate")))

names(data_est_12) <- substring(names(data_est_12), 10)
data_est_12 <- rename(data_est_12, GEOID = 1) 
data_est_12$GEOID <- as.numeric(data_est_12$GEOID)

data_11 <- get_acs(
  geography = "zcta",
  survey = "acs5", 
  cache_table = TRUE, 
  variables = vars,
  year = 2011
  ) %>%
  filter(startsWith(GEOID, "60")) %>% 
  pivot_wider(names_from = variable, values_from = c(estimate, moe)) 

data_est_11 <- data_11 %>% select(c(GEOID, contains("estimate")))

names(data_est_11) <- substring(names(data_est_11), 10)
data_est_11 <- rename(data_est_12, GEOID = 1) 
data_est_11$GEOID <- as.numeric(data_est_11$GEOID)

rm(data_11, data_12, data_13, data_14, data_15, data_16, data_17) #drop dfs we don't need anymore 
```

*Step 3. Read in Raw Shutoffs Data*

```{r}
orders <- read.csv("/Users/emmettmckinney/Dropbox (MIT)/RAship/ESI_Water_Affordability_031120/Chicago-Philly Paper/Chicago/Shutoffs/chicago_shutoffs_2007-2018.csv")
```

*Step 4. Summarize Shutoff Statistics by Zip Code*

```{r}
orders_summary <- orders %>% 
    group_by(zip, year) %>% 
    count(service_order_type) %>% mutate(
    REST = sum(n[service_order_type == "REST"]), #rearranging columnns
    SHUT = sum(n[service_order_type == "SHUT"]),
    IRSO = sum(n[service_order_type == "IRSO"])) %>% 
    select(-service_order_type, -n) %>%
  distinct()
```

*Step 5. Split Up Shutoffs By Year; Join to the ACS-5 survey ending that year then join back together*

```{r}
shutoffs_17 <- filter(orders_summary, year == "2017")
summary_17 <- left_join(shutoffs_17, data_est_17, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years.
rm(shutoffs_17) 

shutoffs_16 <- filter(orders_summary, year == "2016")
summary_16 <- left_join(shutoffs_16, data_est_16, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years.
rm(shutoffs_16) 

shutoffs_15 <- filter(orders_summary, year == "2015")
summary_15 <- left_join(shutoffs_15, data_est_15, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years. 
rm(shutoffs_15) 

shutoffs_14 <- filter(orders_summary, year == "2014")
summary_14 <- left_join(shutoffs_14, data_est_14, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years. 
rm(shutoffs_14) 

shutoffs_13 <- filter(orders_summary, year == "2013")
summary_13 <- left_join(shutoffs_13, data_est_13, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years. 
rm(shutoffs_13) 

shutoffs_12 <- filter(orders_summary, year == "2012")
summary_12 <- left_join(shutoffs_12, data_est_12, copy = FALSE, by = c("zip" = "GEOID"))  #add in census data, lining it up by years. 
rm(shutoffs_12) 

shutoffs_11 <- filter(orders_summary, year == "2011")
summary_11 <- left_join(shutoffs_11, data_est_11, copy = FALSE, by = c("zip" = "GEOID"))
rm(shutoffs_11) #add in census data, lining it up by years. 

shutoffs_summary <- rbind(summary_11, summary_12, summary_13, summary_14, summary_15,summary_16, summary_17) %>% distinct() #join together

rm(summary_11, summary_12, summary_13, summary_14, summary_15,summary_16, summary_17) #get rid of dataframes we don't need
```

*Step 6. Add in new columns to shutoff summary file & write out as .csv*

```{r}
#Add in new columns
shutoffs_summary <- shutoffs_summary %>% 
  mutate(shutoffs_1000p = ((SHUT)/(Total_Pop/1000)),
         shutoffs_HH = (SHUT)/TotalHH,
         Prc_W = White / Total_Pop,
         Prc_Blk = Black / Total_Pop,
         Prc_Hisp = Latino / Total_Pop,
         Prc_W_Own = WhiteOwn/TotalHH,
         Prc_B_Own = BlackOwn/TotalHH,
         Prc_Hisp_Own = HispOwn/TotalHH,
         Prc_Utils_Incl = Utils_Incl / TotalRent,
         Prc_Rnt = TotalRent/TotalHH,
         Prc_Own = TotalOwn/TotalHH,
         Prc_NonW = 1-Prc_W,
         Prc_NonW_Own = 1-Prc_W_Own,
         log_Prc_NonW = log(Prc_NonW),
         log_NonW_Own = log(Prc_NonW_Own),
         log_shut = log(SHUT),
         log_MHI = log(MHI)
        ) %>%
  filter(!is.na(shutoffs_1000p)) %>%
  distinct()

write.csv(shutoffs_summary, file= "/Users/emmettmckinney/Dropbox (MIT)/RAship/ESI_Water_Affordability_031120/Data/WBEZ_Data/Chicago/Data/chicago_zip_year_shutoffs.csv")

shutoffs_zip_allyears <- shutoffs_summary %>% group_by(zip) %>% summarise(
  total_SHUT = sum(SHUT),
  mean_SHUT = mean(SHUT),
  mean_shutoffs_1000p = mean(shutoffs_1000p),
  mean_shutoffs_HH = mean(shutoffs_HH),
  Prc_W = mean(Prc_W),
  Prc_Blk = mean(Prc_Blk),
  Prc_W_Own = mean(Prc_W_Own),
  Prc_B_Own = mean(BlackOwn/TotalHH),
  Prc_Hisp_Own = mean(HispOwn/TotalHH),
  Prc_Utils_Incl = mean(Utils_Incl / TotalRent),
  Prc_Rnt = mean(TotalRent/TotalHH),
  Prc_Own = mean(TotalOwn/TotalHH),
  Prc_NonW = mean(1-Prc_W),
  Prc_NonW_Own = mean(1-Prc_W_Own),
  log_Prc_NonW = mean(log(Prc_NonW)),
  log_NonW_Own = mean(log(Prc_NonW_Own)),
  log_shut = mean(log(SHUT)),
  log_MHI = mean(log(MHI))
) %>% distinct()

write.csv(shutoffs_zip_allyears, file= "/Users/emmettmckinney/Dropbox (MIT)/RAship/ESI_Water_Affordability_031120/Data/WBEZ_Data/Chicago/Data/chicago_zip_allyears.csv")

shutoffs_all_years_geo <- left_join(shutoffs_zip_allyears, geo, by = c("zip" = "GEOID"))
st_write(shutoffs_all_years_geo, "/Users/emmettmckinney/Dropbox (MIT)/RAship/ESI_Water_Affordability_031120/Data/WBEZ_Data/Chicago/Data/chicago_all_years_shutoffs.geojson", delete_dsn = TRUE) 
```

*Step 7. Join .csv to zip code geometry & write out as .geojson*

```{r}
shutoffs_geo <- left_join(shutoffs_summary, geo, by = c("zip" = "GEOID"))
st_write(shutoffs_geo, "chicago_zip_year_shutoffs.geojson", delete_dsn = TRUE) 
```


